Carrier phase and amplitude estimation using pilots and data

### 2012-05-04 paper.tex
Added to complex amp paper.  Started added the statistics

Here, we will describe a new algorithm for computing the least squares estimator that requires only $O(L)$ operations.  We also consider a slightly more general model than is usual, one that allows the utilisation of pilot symbols, i.e., \emph{coherent detection}.  Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

In this section we derive Mackentun's~\cite{Mackenthun1994} least squares estimator.  Mackenthun specifically considered the noncoherent setting, but only minor modifications are required to meaninfully include the pilot symbols, so we also consider the coherent setting here.  For the purpose of anaylsing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  In Section~\ref{sec:line-time-algor} we will show that a full sort is not neccesary, and that the least squares estimator can be implemented in $O(L)$ operations.

Define the sum of squares function
\begin{align}
SS(a, \{d_i, i \in D\}) &= \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*. \label{eq:SS}
\end{align}
Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find that the least squares estimator of $a_0$, satisfies
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\hat{a} = \frac{1}{L}Y$ into~\eqref{eq:SS} we obtain the sum of squares function, conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding sum of squares function $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  We will show that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in the sequence $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k} =\{ y_i f_{ki}^*, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = (e^{j2\pi/M} - 1)$.  In general we have,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k}$ be integer the sequence that minimises $SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata}, as 
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  This descibes a practical implementation of Mackenthun's algorithm.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindicies}$ that, given the sequence $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ of the indices in $D$ as described in~\eqref{eq:sigmasortind}.  Computing the $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all requirse $O(L)$ or less operations.  In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm} \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's least squares estimator}
\dontprintsemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
\For{$i \in D$ \nllabel{alg_loop_setup}}{
$\phi = \angle{y_i}$ \;
$u = \frac{2\pi}{M}\round{\frac{M}{2\pi}\phi} $ \;
$z_i = \phi -  u $ \;
$g_i = y_i e^{-j u}$ \;
}
$\sigma = \operatorname{sortindicies}(z)$ \nllabel{alg_sortindices} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k= 0$ \emph{\textbf{to}} $(M+1)\abs{D}-1$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/M} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = g_{\sigma(k)} e^{j2\pi/M} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

%Let $f$ be a function maping $\reals$ to $\reals^N$ defined so that the $i$th element of $f(\theta)$ is $\round{\frac{M}{2\pi}( \phi_i - \theta )}$.  Let $\phi$ be the vector of length $N$ with element $[\phi_1, \dots, \phi_N]$ and define the $\round{\cdots}$ to operate elementwise on vectors, then
%\[
%f(\theta) = \round{\frac{M}{2\pi}( \phi - \theta )}.
%\]
%The function $f(\theta)$ is piecewise continuous.

>>>>

Here, consider a slightly more general model than is usual, one that allows the utilisation of pilot symbols, i.e., \emph{coherent detection}.  Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

In this section we derive Mackentun's~\cite{Mackenthun1994} least squares estimator.  Mackenthun specifically considered the noncoherent setting, but only minor modifications are required to meaninfully include the pilot symbols, so we also consider the coherent setting here.  For the purpose of anaylsing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  %In Section~\ref{sec:line-time-algor} we will show that a full sort is not neccesary, and that the least squares estimator can be implemented in $O(L)$ operations.

Define the sum of squares function
\begin{align}
SS(a, \{d_i, i \in D\}) &= \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*. \label{eq:SS}
\end{align}
Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find that the least squares estimator of $a_0$, satisfies
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\hat{a} = \frac{1}{L}Y$ into~\eqref{eq:SS} we obtain the sum of squares function, conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding sum of squares function $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  We will show that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in the sequence $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k} =\{ y_i f_{ki}^*, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = (e^{j2\pi/M} - 1)$.  In general we have,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k}$ be the integer that minimises $SS(f_k)$, i.e. $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata}, as 
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindicies}$ that, given the sequence $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ of the indices in $D$ as described in~\eqref{eq:sigmasortind}.  Computing the $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all requirse $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm} \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's least squares estimator}
\DontPrintSemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
\For{$i \in D$ \nllabel{alg_loop_setup}}{
$\phi = \angle{y_i}$ \;
$u = \frac{2\pi}{M}\round{\frac{M}{2\pi}\phi} $ \;
$z_i = \phi -  u $ \;
$g_i = y_i e^{-j u}$ \;
}
$\sigma = \operatorname{sortindicies}(z)$ \nllabel{alg_sortindices} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k= 0$ \emph{\textbf{to}} $(M+1)\abs{D}-1$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/M} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = g_{\sigma(k)} e^{j2\pi/M} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

BLERG $\hat{a}$ is a biased estimator, particularly when the SNR is low, we show firs that $\hat{\theta}$ is strongly consistent, we are then able to analyse the behaviour of $\hat{a}$.

Again let $a = \rho e^{j\theta}$ and write the sum of squares function  
\[
SS(\rho, \theta, \{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{y_i}^2 - \rho y_i e^{-j\theta} s_i^* - \rho y_i e^{-\theta} s_i + \rho^2.
\]
Differentiating with respect to $\rho$ and setting the resulting expression to zero gives
\[
\rho = \frac{Z + Z^*}{2L}
\]
where $Z = \sum_{i\in P \cup D}y_i e^{-j\theta} s_i^*$.  Observe that $Z$ is really a function of $\theta$ and $\{d_i, i \in D\}$, but we have supressed the arguments for notational clarity.  We can now write
\[
SS(\rho, \theta, \{d_i, i \in D\}) = A - \rho Z - \rho Z^* + L \rho^2
\]
and substituting the above expression for $\rho$ into this gives the sum of squares function conditioned upon minimisation with respect to $\rho$,
\[
SS(\theta, \{d_i, i \in D\}) = A - \frac{1}{L}\abs{Z}^2.
\]
Substituting~(\ref{eq:hatdfinxtheta}) into this gives the sum of squares function conditioned upon minimisation with respect to $\rho$ and the data sequence $\{d_i, i \in D\}$,
\[
SS(\theta) = A - \frac{1}{L}\abs{Z(\theta)}^2
\]
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta).
\]
We have reused $Z$ here, but this should not cause confusion as $Z(\theta)$ and $Z(\theta, \{d_i, i \in D\})$ are easily told apart from there arguments.  From now on, when we right $Z$ without arguments, we will always be refering to $Z(\theta)$.

We are interested in analysing $\hat{\theta}$ the maximiser of $\abs{Z(\theta)}^2$.  Put $\lambda_L = \fracpart{\theta_0 - \hat{\theta}}$.  We want to show that $\lambda_L$ converges to zero as the number of transmitted symbols $L$ increases to $\infty$ where $\fracpart{x}$ denotes $x$ taken `modulo $2\pi$' into the interval $[-\pi, \pi)$, that is
\[
\fracpart{x} = x - 2\pi\round{\frac{x}{2\pi}}.
\]
Under the assumption that the noise $\{w_i\}$ is circularly symetric then
\[
y_i = a_0 s_i + w_i = \rho_j e^{j ( \theta_0 + \Phi_i + k_i) }
\] 
where $k_i = \angle s_i$ and the $\Phi_i$ are identically distributed.  The $\Phi_i$ can be identified as \emph{circular random variables}~\cite{Mardia_directional_statistics,Fisher1993,McKilliam_mean_dir_est_sq_arc_length2010}.  Recalling the definition of $\hat{d}_i(\theta)$ and $\hat{u}_i(\theta)$ from~\eqref{eq:hatdfinxtheta}, we have
\[
\hat{u}_i(\theta) = \round{\frac{M}{2\pi}(\angle{y_i} - \theta)} = \round{\frac{M}{2\pi}(\lambda + \Phi_i + k_i)}
\]
where $\lambda = \fracpart{\theta_0 - \theta}$.  So,
\begin{align*}
 y_i e^{-j\theta} \hat{d}_i(\theta)^* &= \rho_j \exp{j \left(\lambda + \Phi_i + k_i - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\lambda + \Phi_i + k_i)}\right)} \\
&= \rho_j \exp{j \left(\lambda + \Phi_i - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\lambda + \Phi_i) }\right)}
\end{align*}
when $i \in D$ and
\[
y_i e^{-j\theta} p_i^* = \rho_j \exp j \left(\lambda + \Phi_i )
\]
when $i \in P$.  Now we can write
\[
Z(\theta) = \sum_{i \in P} \rho_j e^{j(\lambda + \Phi_i)} + \sum_{i \in D}  \rho_j \exp{j \left( \lambda + \Phi_i - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\lambda + \Phi_i)}\right)}.
\]
Put $S_L(\lambda) = \frac{1}{L}\abs{Z(\theta)}^2$ and put $\lambda_L = \fracpart{\theta_0 - \hat{\theta}}$.  Then $\lambda_L$ is the maximiser of $S_L(\lambda)$.  We will show that $\lambda_L$ converges almost surely to zero as $L \rightarrow \infty$.  From this, the proof of strong consistency follows.

BLERG  Show $E S_L$ is maximised uniquely at 0, proof follows.

We now prove the asymptotics normality of the estimator.  At the same time we are able to describe the distribution of the amplitude estimator $\hat{\rho}$.  BLERG, first show that it is differentiable at the truth.
